The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2010-01-18 script/de/chapters/primes.tex
Script[de+en]: Small updates in primes chapter.

  \item \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
     Es existiert eine reelle Zahl a, so dass die Funktion
     $f: {\bf N} \rightarrow {\mathbb Z}$ mit $n \mapsto a^{3^n}$
     für alle $n$ nur Primzahlenwerte annimmt 
     (siehe \cite[S. 82]{pr:Padberg1996}).
     Leider macht die Bestimmung von $a$ Probleme 
     (siehe \hyperlink{link-Primzahlfunktion-base-a-Offen}{unten}).

@@

     Die längste Primzahlfolge, die man bisher kennt, hat 25 Glieder.
     In der Tabelle~\ref{Smallest AP-k with minimal difference}
     sind die längsten bekannten arithmetischen
     Primzahlfolgen mit minimaler Distanz\footnote{%
        Dagegen sind in \url{http://en.wikipedia.org/wiki/Primes_in_arithmetic_progression}
        die "Largest known AP-k" aufgelistet hat, also dass das letzte
        Folgenelement eine möglichst große Primzahl ist.
        Hier sind die Folgen aufgelistet, die die kleinsten bekannten
        Differenzen haben -- für eine gegebene Folgenlänge.
     }
     aufgelistet.

@@

     Eigentlich hatten Green und Tao nur beweisen wollen, dass es unendlich
     viele arithmetische Primzahlfolgen mit vier Gliedern gibt. Dazu
     betrachteten sie Mengen, die neben Primzahlen auch 
     Beinaheprimzahlen\index{Beinaheprimzahl}\index{Primzahl!Beinaheprimzahl} 
     enthielten. Das sind Zahlen, die nur wenige Teiler haben - beispielsweise
     die Halbprimzahlen\index{Halbprimzahl}\index{Primzahl!Halbprimzahl}
     \index{Zahlen!semiprime}, die Produkte aus genau zwei Primzahlen sind.
     Dadurch konnten die beiden Mathematiker ihre Arbeit wesentlich
     erleichtern, denn über Beinaheprimzahlen gab es schon zahlreiche
     nützliche Theoreme. Schließlich erkannten sie, dass ihr Verfahren viel
     mächtiger ist, als sie selbst angenommen hatten, und sie bewiesen damit
     die Hardysche Vermutung.

@@

     \begin{table}[ht]
     \hspace*{1cm}\begin{tabular}{|r|r|}
     \hline
      k &                k\# \\ \hline
      2 &                 2 \\
      3 &                 6 \\
      5 &                30 \\
      7 &               210 \\
     11 &             2.310 \\
     13 &            30.030 \\
     17 &           510.510 \\
     19 &         9.699.690 \\
     23 &       223.092.870 \\
      \hline
     \end{tabular}
     \caption{Produkte der ersten Primzahlen $<= k$ (genannt k Primorial oder $k\#$)}
     \end{table}

@@

\begin{itemize}
\item \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
    Der Beweis \hyperlink{link-Primzahlfunktion-base-a-Proof}{oben} zu der
    Funktion  $f: \: N \rightarrow Z$ mit $n \mapsto a^{3^n}$
    garantiert nur die Existenz einer
    solchen Zahl $a$.  Wie kann diese Zahl $a$ bestimmt werden, und wird sie
    einen Wert haben, so dass die Funktion auch von praktischem Interesse ist?

>>>>

  \item % \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
        \label{link-Primzahlfunktion-base-a-Proof}
     Es existiert eine reelle Zahl a, so dass die Funktion
     $f: {\bf N} \rightarrow {\mathbb Z}$ mit $n \mapsto \lfloor a^{3^n}\rfloor$
     für alle $n$ nur Primzahlenwerte annimmt\footnote{%
     Die Gaussklammer \index{Gaussklammer} $\lfloor x \rfloor $ der reellwertigen Zahl $x$
     ist definiert als: $\lfloor x \rfloor $ ist die größte ganze Zahl kleiner oder gleich $x$.
     } (siehe \cite[S. 82]{pr:Padberg1996}).
     Leider macht die Bestimmung von $a$ Probleme 
     (siehe Kapitel \ref{link-Primzahlfunktion-base-a-Offen}).

% \label{zahlentyp_mersenne}
% \hypertarget{GIMPS-project}{}
% \hyperlink{GIMPS-project}{GIMPS-Projekt} (Kapitel~\ref{zahlentyp_mersenne}),

@@

     Die längste Primzahlfolge, die man bisher kennt, hat 25 Glieder.
     In der Tabelle~\ref{Smallest AP-k with minimal difference}
     sind die längsten bekannten arithmetischen
     Primzahlfolgen mit minimaler Distanz\footnote{%
        Dagegen sind in \url{http://en.wikipedia.org/wiki/Primes_in_arithmetic_progression}
        die \glqq Largest known AP-k\grqq~aufgelistet. Also ist dort das letzte
        Folgenelement eine möglichst große Primzahl.
        Hier jedoch sind die Folgen aufgelistet, die die kleinsten bekannten
        Differenzen haben -- für eine gegebene Folgenlänge.
     }
     aufgelistet.

@@

     Eigentlich hatten Green und Tao nur beweisen wollen, dass es unendlich
     viele arithmetische Primzahlfolgen mit vier Gliedern gibt. Dazu
     betrachteten sie Mengen, die neben Primzahlen auch 
     Beinaheprimzahlen\index{Beinaheprimzahl}\index{Primzahl!Beinaheprimzahl} 
     enthielten. Das sind Zahlen, die nur wenige Teiler haben - beispielsweise
     die Halbprimzahlen%
     \index{Halbprimzahl}\index{Primzahl!Halbprimzahl}%
     \index{Zahlen!semiprime}\index{Zahlen!Halbprimzahl},
     die Produkte aus genau zwei Primzahlen sind.
     Dadurch konnten die beiden Mathematiker ihre Arbeit wesentlich
     erleichtern, denn über Beinaheprimzahlen gab es schon zahlreiche
     nützliche Theoreme. Schließlich erkannten sie, dass ihr Verfahren viel
     mächtiger ist, als sie selbst angenommen hatten, und sie bewiesen damit
     die Hardysche Vermutung.

@@

     \begin{table}[ht]
     \begin{center}
     \hspace*{1cm}\begin{tabular}{|r|r|}
     \hline
      k &                k\# \\ \hline
      2 &                 2 \\
      3 &                 6 \\
      5 &                30 \\
      7 &               210 \\
     11 &             2.310 \\
     13 &            30.030 \\
     17 &           510.510 \\
     19 &         9.699.690 \\
     23 &       223.092.870 \\
      \hline
     \end{tabular}
     \caption{Produkte der ersten Primzahlen $<= k$ (genannt k Primorial oder $k\#$)}
     \end{center}
     \end{table}

@@

\begin{itemize}
\item % \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
      \label{link-Primzahlfunktion-base-a-Offen}
    Der in Kapitel \ref{link-Primzahlfunktion-base-a-Proof} erwähnte Beweis zu der
    Funktion  $f: \: N \rightarrow Z$ mit $n \mapsto \lfloor a^{3^n}\rfloor$
    garantiert nur die Existenz einer
    solchen Zahl $a$.  Wie kann diese Zahl $a$ bestimmt werden, und wird sie
    einen Wert haben, so dass die Funktion auch von praktischem Interesse ist?

### 2010-01-18 script/en/chapters/primes.tex
Script[de+en]: Small updates in primes chapter.

  \item  \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
      There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{pr:Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ 
      (\hyperlink{link-Primzahlfunktion-base-a-Offen}{see below}).

@@

     The longest arithmetic prime number sequence known today contains 
     25 elements.
     The table~\ref{Smallest AP-k with minimal difference}
     lists the longest currently known arithmetic prime number sequences
     with minimal difference\footnote{%
        In the opposite, in \url{http://en.wikipedia.org/wiki/Primes_in_arithmetic_progression}
        are the "largest known AP-k" listed. Therefore, the last sequence element
        is a prime as big as possible. However, here are the sequences listed
        which have the smallest known difference for a given length.
     }.

@@

     Green and Tao intended to proof that there are infinitely many arithmetic
     sequences of length four. For this they considered sets of numbers 
     consisting of primes and so called 
     ``near primes''\index{Near prime}\index{Prime number!near prime}.
     These are numbers with a small set of divisors like numbers which are
     the product of two primes - these numbers are called 
     ``half primes''\index{Half prime}\index{Prime number!half prime}
     \index{Prime number!semi prime}\index{Prime number!bi prime}
     \index{Number!semi prime}\index{Number!bi prime}.
     Thus they managed to considerably simplify their work because about near
     primes there already existed a lot of useful theorems.  Finally they
     discovered that the results of their theorem were far more reaching than
     they had assumed and so they were able to prove Hardy's conjecture.

@@

     If someone wants to discover such sequences he should consider the
     following thought.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 results as the product of all primes smaller than the
     length of the sequence. So its the product of all primes smaller than 6:
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     It is $ 10\# = 7\# = 2 ∑ 3 ∑ 5 ∑ 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $ 15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030 $.

@@

     \begin{table}[ht]
     \hspace*{1cm}\begin{tabular}{|r|r|}
     \hline
      k &                k\# \\ \hline
      2 &                 2 \\
      3 &                 6 \\
      5 &                30 \\
      7 &               210 \\
     11 &             2.310 \\
     13 &            30.030 \\
     17 &           510.510 \\
     19 &         9.699.690 \\
     23 &       223.092.870 \\
      \hline
     \end{tabular}
     \caption{Products of the first primes $<= k$ (called k primorial or $k\#$)}
     \end{table}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Unproven statements / conjectures about primes}
\begin{itemize}
  \item Christian Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.
     Computers have verified\footnote{%
     It is generally accepted today, that the \textbf{Goldbach Conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: see\\
     \url{http://www.ieeta.pt/~tos/goldbach.html},\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \url{http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to
     $15*10^{17}$ (as at July 2009) but no general proof has yet been
     found\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.\\
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.\\
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers. }.

@@

\begin{itemize}
\item \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
    The \hyperlink{link-Primzahlfunktion-base-a-Proof}{above} proof of the
    function $f: \: N \rightarrow Z$ with $n \mapsto a^{3^n}$  
    only guarantees the existence of such a number $a$.  How can we
    determine this number $a$ and will it have a value, making the 
    function also of some practical interest?

>>>>

  \item % \hypertarget{link-Primzahlfunktion-base-a-Proof}{}
        \label{link-Primzahlfunktion-base-a-Proof}
      There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto \lfloor a^{3^n}\rfloor$ 
      only delivers primes\footnote{%
      The Gauss\index{Gauss bracket} bracket  $\lfloor x \rfloor $ of a real number $x$
      is defined via: $\lfloor x \rfloor $ is the next integer less or equal $x$.
      } for all $n$  (see \cite[p. 82]{pr:Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ 
      (see chapter \ref{link-Primzahlfunktion-base-a-Offen}).

@@

     The longest arithmetic prime number sequence known today contains 
     25 elements.
     The table~\ref{Smallest AP-k with minimal difference}
     lists the longest currently known arithmetic prime number sequences
     with minimal difference\footnote{%
        In the opposite, in \url{http://en.wikipedia.org/wiki/Primes_in_arithmetic_progression}
        are the "largest known AP-k" listed. Therefore, there the last sequence element
        is a prime as big as possible. However, here are the sequences listed
        which have the smallest known difference for a given length.
     }.

@@

     Green and Tao intended to proof that there are infinitely many arithmetic
     sequences of length four. For this they considered sets of numbers 
     consisting of primes and so called 
     ``near primes''\index{Near prime}\index{Prime number!near prime}.
     These are numbers with a small set of divisors like numbers which are
     the product of two primes - these numbers are called 
     ``half primes''%
     \index{Half prime}\index{Prime number!half prime}%
     \index{Prime number!semi prime}\index{Prime number!bi prime}%
     \index{Number!semi prime}\index{Number!bi prime}.
     Thus they managed to considerably simplify their work because about near
     primes there already existed a lot of useful theorems.  Finally they
     discovered that the results of their theorem were far more reaching than
     they had assumed and so they were able to prove Hardy's conjecture.

@@

     If someone wants to discover such sequences he should consider the
     following thought.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 results as the product of all primes smaller than the
     length of the sequence. So its the product of all primes smaller than 6:
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     It is $ 10\# = 7\# = 2 ∑ 3 ∑ 5 ∑ 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030$.

@@

     \begin{table}[ht]
     \begin{center}
     \hspace*{1cm}\begin{tabular}{|r|r|}
     \hline
      k &                k\# \\ \hline
      2 &                 2 \\
      3 &                 6 \\
      5 &                30 \\
      7 &               210 \\
     11 &             2,310 \\
     13 &            30,030 \\
     17 &           510,510 \\
     19 &         9,699,690 \\
     23 &       223,092,870 \\
      \hline
     \end{tabular}
     \caption{Products of the first primes $<= k$ (called k primorial or $k\#$)}
     \end{center}
     \end{table}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{Unproven statements / conjectures about primes}
\begin{itemize}
  \item Christian Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.
     Computers have verified\footnote{%
     It is generally accepted today, that the \textbf{Goldbach Conjecture} is true, 
     i.~e.\ valid for all even natural numbers greater than $2$.
     In 1999, mathematician J\"org Richstein\index{Richstein 1999}
     from the computer sciences institute at the University of Giessen,
     studied even numbers up to 400 billion ($4*10^{14}$) and found no
     contradictory example (\cite{pr:Richstein1999}).\\
     In the meantime further progress was made: See\\
     \url{http://www.ieeta.pt/~tos/goldbach.html},\\
     \url{http://en.wikipedia.org/wiki/Goldbach's\_conjecture},\\
     \url{http://primes.utm.edu/glossary/page.php/GoldbachConjecture.html}.\\      
     Nevertheless, this does not provide us with general proof.\\
     The fact is that despite all efforts, Goldbach's conjecture has to 
     date not been proven. This leads one to believe that since the pioneer
     work of the Austrian mathematician Kurt G\"odel \index{G\""odel, Kurt}
     is well-known, not every true mathematical theorem is provable (see
     \url{http://www.mathematik.ch/mathematiker/goedel.html}).
     Perhaps Goldbach's conjecture was correct, but in any case the proof
     will never be found. Conversely, that will presumably also remain 
     unproven.}
     the Goldbach conjecture for all even numbers up to
     $15*10^{17}$ (as at July 2009) but no general proof has yet been
     found\footnote{%
     The English publisher {\em Faber} and the American publisher 
     {\em Bloomsbury} issued in 2000 the 1992 published book 
     ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
     It's the story of an old maths professor who fails to prove a more 
     than 250 year old puzzle. To boost the sales figures the English and
     American publishers have offered a prize of 1 million USD, if someone
     can prove the conjecture -- which should be published by 2004 in a 
     well-known mathematical journal.\\
     Surprisingly only British and American citizens are allowed to 
     participate.\\
     The theorem which has come closest so far to Goldbach's conjecture
     was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
     to understand:
     Each even integer greater than $2$ is the sum of one prime and of 
     the product of two primes. E.g.: $20=5+3*5.$\\
     Most of the research about the Goldbach conjecture is collected in 
     the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World 
     scientific Series in Pure Maths, Vol. 4.\\
     Especially this conjecture makes it clear, that even today we do not
     have a complete understanding of the deeper connections between 
     addition and multiplication of natural numbers. }.

@@

\begin{itemize}
\item % \hypertarget{link-Primzahlfunktion-base-a-Offen}{}
      \label{link-Primzahlfunktion-base-a-Offen}
    The proof (mentioned above in chapter \ref{link-Primzahlfunktion-base-a-Proof}) of the
    function $f: \: N \rightarrow Z$ with $n \mapsto \lfloor a^{3^n}\rfloor$  
    only guarantees the existence of such a number $a$.  How can we
    determine this number $a$ and will it have a value, making the 
    function also of some practical interest?
